---
title: "Reproducible Research"
author: "Boris Shminke"
date: "17.02.2015"
output:
  html_document:
    keep_md: true
---

## Synopsis

In this report we use Strom Events data from National Weather Service to find events most harmful with respect to population health and events which have the greatest economic consequences.

## Data Acquisition

We download data if needed and read it from *bzip2*-file.
This operation is cached because of it's quite time-consuming.

```{r, cache=TRUE}
bzfilename <- "repdata-data-StormData.csv.bz2"
if(!file.exists(bzfilename)) {
  download.file(
    "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
    bzfilename, "curl")
}

data <- read.csv(bzfile(bzfilename))
```

## Data Processing

Here we use `dplyr` package for simple data analysis.
We define harm to population health as total number of fatalities and injuries.
We define economic consequences as total damage to property and crops.

```{r}
library(dplyr)
worst_h <- group_by(data, EVTYPE) %>% summarise(health=sum(FATALITIES)+sum(INJURIES)) %>% arrange(desc(health))
worst_e <- group_by(data, EVTYPE) %>% summarise(economy=sum(PROPDMG)+sum(CROPDMG)) %>% arrange(desc(economy))
```

## Results

As we can see from data processing section, most harmful events with respect to population health are (from the most harmful to less): `r paste(head(worst_h)$EVTYPE, collapse = ", ")`.
And the following events have the greatest economic consequences  (from the most harmful to less): `r paste(head(worst_e)$EVTYPE, collapse = ", ")`.

```{r}
  par(mar = c(10,4,4,2) + 0.1)
  barplot(head(worst_h)$health, names.arg = head(worst_h)$EVTYPE, las=3,
          main="Fatalities and injuries by event type")
```

```{r}
  par(mar = c(10,4,4,2) + 0.1)
  barplot(head(worst_e)$economy/1000000, names.arg = head(worst_e)$EVTYPE,
          las=3,
          main="Property and crop damage in millions of USD by event type")
```